Choosing TTC Subway Service Provides the Most Time Stability during Weekday Rush-Hours

### 2024-09-24 outputs/paper/paper.qmd
editted code style in paper and re-rendered

```{r}
#| include: false
#| warning: false
#| message: false
# load librarys and datasets
library(tidyverse)
library(dplyr)
library(knitr)
library(ggplot2)
library(hms)
library(patchwork)

@@

# Introduction
Public transportation serves as the transportation backbone for city residents, facilitating the daily commute of millions of people. For office workers, students, and the elderly, selecting the appropriate TTC service is crucial in maintaining a reliable and efficient schedule. Despite the importance of this choice, there has been limited effort to evaluate the reliability of different TTC services to guide commuters’ choices.

@@

# Data {#sec-data}
Where the data comes from
- All data used for and throughout the paper are extracted from Open Data Toronto Portal [@OpenDataToronto]. TTC Bus Delay data [@Bus], TTC Streetcar Delay [@Streetcar], TTC Subway Delay [@Subway] and TTC Subway Delay Codes [@Subway] are retrieved to compare the delay magnitudes of different services. Data is cleaned and analyzed using the open source statistical programming language R [@citeR] and supporting packages tidyverse [@tidyverse], janitor [@janitor], dplyr [@dplyr], hms [@hms], ggplot2 [@ggplot2], patchwork [@patchwork], and knitr [@knitr].

For all delay dataset, only data in 2023 were chosen. There are three main reasons for this choice:

1. 2023 data is the most recent full-year data and offers the best reference for current (2024) delay patterns. 
2. COVID-19 impacted TTC operating capacity and styles in other years close to 2024. 
3. data from earlier years (before 2020) were structured very differently.

@@

## TTC Bus Delay {#sec-data-bus}
source of data (brief)
structure of data & important attributes
why are these attributes kept for analysis (measurement)

@@

## TTC Streetcar Delay {#sec-data-streetcar}
source of data (brief)
structure of data & important attributes
why are these attributes kept for analysis (measurement)

@@

## TTC Subway Delay {#sec-data-subway}
source of data (brief)

### Delay Code {#sec-data-code}
purpose of extra dataset
how it is related to subway_delay
```{r}
#| label: tbl-subway-code
#| tbl-cap: Sample of TTC Subway Delay Code
#| echo: false
#| message: false

@@

### Delay Data
structure of data & important attributes
why are these attributes kept for analysis (measurement)
mention "code" column
```{r}
#| label: tbl-subway-delay
#| tbl-cap: Sample of Cleaned TTC Subway Delay Dataset
#| echo: false
#| message: false

@@

# Results {#sec-result}

@@

## Most Delayed during Weekday
Define Rush Hours:
  Rush Hours: 
  •	Morning Rush-Hour: Approximately 7:00 AM to 9:00 AM
	•	Evening Rush-Hour: Approximately 4:00 PM to 7:00 PM

@@

## Most Delayed during Rush-Hours
```{r}
#| echo: false
#| message: false
# Setup
bus_delay_rush <- read.csv(here::here("inputs/data/cleaned_data/bus/bus_delay_rush.csv"))
streetcar_delay_rush <- read.csv(here::here("inputs/data/cleaned_data/streetcar/streetcar_delay_rush.csv"))
subway_delay_rush <- read.csv(here::here("inputs/data/cleaned_data/subway/subway_delay_rush.csv"))
```

@@

## Most Popular Delay Reasons
- 1 table for all three services
- separate into rush & off-peak hours
```{r}
#| label: tbl-reason
#| tbl-cap: Top 5 Most Popular Reason for Delay
#| echo: false
#| message: false

@@

## Most Popular Delayed Lines
- 1 table for all three services
- separate into rush & off-peak hours
```{r}
#| label: tbl-lines
#| tbl-cap: Most delayed Lines for each TTC Service
#| echo: false
#| message: false

@@

## Correlation between Delay and Direction of Travel
- Financial District
- Midtown Direction

## TTC Service Amenities
- The PATH (for financial district)
- AC in Summer & Winter

## Errors in analysis, Limitations and Next Steps

>>>>

```{r}
#| include: false
#| warning: false
#| message: false
# load libraries and datasets
library(tidyverse)
library(dplyr)
library(knitr)
library(ggplot2)
library(hms)
library(patchwork)

@@

# Introduction

Public transportation serves as the transportation backbone for city  residents, facilitating the daily commute of millions of people. For office workers, students, and the elderly, selecting the appropriate TTC service is crucial in maintaining a reliable and efficient schedule. Despite the importance of this choice, there has been limited effort to evaluate the reliability of different TTC services to guide commuters’choices.

@@

# Data {#sec-data}

Where the data comes from - All data used for and throughout the paper are extracted from Open Data Toronto Portal [@OpenDataToronto]. TTC Bus Delay data [@Bus], TTC Streetcar Delay [@Streetcar], TTC Subway Delay [@Subway] and TTC Subway Delay Codes [@Subway] are retrieved to compare the delay magnitudes of different services. Data is cleaned and analyzed using the open source statistical programming language R [@citeR] and supporting packages tidyverse [@tidyverse], janitor [@janitor], dplyr [@dplyr], hms [@hms], ggplot2 [@ggplot2], patchwork [@patchwork], and knitr [@knitr].
\
\
For all delay dataset, only data in 2023 were chosen. There are three
main reasons for this choice:

**1.** 2023 data is the most recent full-year data and offers the best reference for current (2024) delay patterns.

**2.** COVID-19 impacted TTC operating capacity and styles in other years close to 2024.

**3.** data from earlier years (before 2020) were structured very differently.

@@

## TTC Bus Delay {#sec-data-bus}

source of data (brief) structure of data & important attributes

why are these attributes kept for analysis (measurement)

@@

## TTC Streetcar Delay {#sec-data-streetcar}

source of data (brief) structure of data & important attributes

why are these attributes kept for analysis (measurement)

@@

## TTC Subway Delay {#sec-data-subway}

source of data (brief)

### Delay Code {#sec-data-code}

purpose of extra dataset how it is related to subway_delay

```{r}
#| label: tbl-subway-code
#| tbl-cap: Sample of TTC Subway Delay Code
#| echo: false
#| message: false

@@

### Delay Data

structure of data & important attributes

why are these attributes kept for analysis (measurement) mention "code" column

```{r}
#| label: tbl-subway-delay
#| tbl-cap: Sample of Cleaned TTC Subway Delay Dataset
#| echo: false
#| message: false

@@

# Results {#sec-results}

@@

## Most Delayed during Weekday

Define Rush Hours: Rush Hours: 
• Morning Rush-Hour: Approximately 7:00 AM to 9:00 AM 
• Evening Rush-Hour: Approximately 4:00 PM to 7:00 PM

@@

## Most Delayed during Rush-Hours

```{r}
#| echo: false
#| message: false
# Setup
bus_delay_rush <- read.csv(here::here("inputs/data/cleaned_data/bus/bus_delay_rush.csv"))
streetcar_delay_rush <- read.csv(here::here("inputs/data/cleaned_data/streetcar/streetcar_delay_rush.csv"))
subway_delay_rush <- read.csv(here::here("inputs/data/cleaned_data/subway/subway_delay_rush.csv"))
```

@@

## Most Popular Delay Reasons

-   1 table for all three services
-   separate into rush & off-peak hours

```{r}
#| label: tbl-reason
#| tbl-cap: Top 5 Most Popular Reason for Delay
#| echo: false
#| message: false

@@

## Most Popular Delayed Lines

-   1 table for all three services
-   separate into rush & off-peak hours

```{r}
#| label: tbl-lines
#| tbl-cap: Most delayed Lines for each TTC Service
#| echo: false
#| message: false

@@

## Importance of TTC Services
- capacity of TTC daily TTC service capacities
- rush-hour traffic

## TTC Service Amenities

- The PATH (for financial district)
- AC in Summer & Winter

## Errors in analysis, Limitations and Next Steps
- only included 2023 data, should include more years for fairer inspection
- unable to distinguish between on-time rate for different district
- next step: explore GO services (GO trains)
